Correctly Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
Moved figure.
Some punctuation fixes.
Punctuation.
Some fixes in Section 5

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

\begin{theorem}
Timestamps are a complete abstraction
\end{theorem}
The proof follows from the galois connection that, in one direction,
sorts all the stores in $S$ to form $\Sigma$, and translates stores in
$S$ to their distance from the end of $\Sigma$. In the other
direction, timestamps are replaced by the stores they point to.

@@

A taste of the changes to the reduction relation is as follows:

\begin{align*}
\dmachstep &\subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) \\
\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

\spchoice
{
\begin{theorem}[Soundness]
  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq
  \mastate$ then there exists a $\mastate'$ such that $\mastate
  \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
\end{theorem}
Here $\sqsubseteq$ is straightforward - the LHS store must be
contained in the RHS store, and if values occur in the states, forcing
the LHS value must be a subset of forcing the corresponding RHS value.
The proof is by cases on $\mstate \machstep \mstate'$.}
{
\begin{theorem}[Completeness]For all $\mexp$,
 $\traces_{\text{Lazy}}(\mexp)$ is a complete abstraction of $\traces_{\text{ISWIM}}(\mexp)$.
\end{theorem}
We have a statement about traces because we need induction to show no
cruft values are in superposition.  The induction hypothesis tells us
that there are non-lazy traces that lead to all the values in
superposition, so when we take a lazy step, we are taking several
non-lazy steps, and we stay in sync. The other direction we just
collapse the superposition in each possibility to construct the
non-lazy traces.}

>>>>

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics;
 \item lazy non-determinism; and
 \item abstract compilation.
\end{enumerate}

@@

\begin{theorem}
Timestamps are a complete abstraction.
\end{theorem}
The proof follows from the Galois connection that, in one direction,
sorts all the stores in $S$ to form $\Sigma$, and translates stores in
$S$ to their distance from the end of $\Sigma$. In the other
direction, timestamps are replaced by the stores they point to.

@@

A taste of the changes to the reduction relation is as follows:
\begin{equation*}
\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
\end{equation*}
\begin{align*}
\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

\spchoice
{
\begin{theorem}[Soundness]
  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq
  \mastate$ then there exists a $\mastate'$ such that $\mastate
  \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
\end{theorem}
Here $\sqsubseteq$ is straightforward - the left-hand side store must be
contained in the right-hand-side store, and if values occur in the states, forcing
the left-hand-side value must be a subset of forcing the corresponding right-hand-side value.
The proof is by cases on $\mstate \machstep \mstate'$.}
{
\begin{theorem}[Completeness]For all $\mexp$,
 $\traces_{\text{Lazy}}(\mexp)$ is a complete abstraction of $\traces_{\text{ISWIM}}(\mexp)$.
\end{theorem}
We have a statement about traces because we need induction to show no
cruft values are in superposition.  The induction hypothesis tells us
that there are non-lazy traces that lead to all the values in
superposition, so when we take a lazy step, we are taking several
non-lazy steps, and we stay in sync. The other direction we just
collapse the superposition in each possibility to construct the
non-lazy traces.}
